# OBITools V4 Tutorial

Here is a short tutorial on how to analyze DNA metabarcoding data produced on Illumina sequencers using:

- the OBITools
- some basic Unix commands

## Wolves’ diet based on DNA metabarcoding

The data used in this tutorial correspond to the analysis of four wolf scats, using the protocol published in @Shehzad2012-pn for assessing carnivore diet. After extracting DNA from the faeces, the DNA amplifications were carried out using the primers `TTAGATACCCCACTATGC` and `TAGAACAGGCTCCTCTAG` amplifiying the *12S-V5* region [@Riaz2011-gn], together with a wolf blocking oligonucleotide.

The complete data set can be downloaded here: [the tutorial dataset](wolf_diet.tgz) 

Once the data file is downloaded, using a UNIX terminal unarchive the data from the `tgz` file.

```{bash untar_data}
#| output: false
tar zxvf wolf_diet.tgz
```

That command create a new directory named `wolf_data` containing every required data files:

-   `fastq <fastq>` files resulting of aGA IIx (Illumina) paired-end (2 x 108 bp) 
    sequencing assay of DNA extracted and amplified from four wolf faeces:

    -   `wolf_F.fastq`
    -   `wolf_R.fastq`

-   the file describing the primers and tags used for all samples
    sequenced:

    -   `wolf_diet_ngsfilter.txt` The tags correspond to short and
        specific sequences added on the 5\' end of each primer to
        distinguish the different samples

-   the file containing the reference database in a fasta format:

    -   `db_v05_r117.fasta` This reference database has been extracted
        from the release 117 of EMBL using `obipcr`

```{bash true_mk_directory}
#| output: false
#| echo: false
#| error: true
#| 
if [[ ! -d results ]] ; then
  mkdir results
fi
```

To not mix raw data and processed data a new directory called `results` is created.

```{bash mk_directory}
#| output: false
#| eval: false
mkdir results
```
 
## Step by step analysis

### Recover full sequence reads from forward and reverse partial reads

When using the result of a paired-end sequencing assay with supposedly
overlapping forward and reverse reads, the first step is to recover the
assembled sequence.

The forward and reverse reads of the same fragment are *at the same line
position* in the two fastq files obtained after sequencing. Based on
these two files, the assembly of the forward and reverse reads is done
with the `obipairing` utility that aligns the two reads and returns the
reconstructed sequence.

In our case, the command is:

```{bash pairing}
#| output: false

obipairing --min-identity=0.8 \
           --min-overlap=10 \
           -F wolf_data/wolf_F.fastq \
           -R wolf_data/wolf_R.fastq \
           > results/wolf.fastq 
```

The `--min-identity` and `--min-overlap` options allow
discarding sequences with low alignment quality. If after the aligment,
the overlaping parts of the reads is shorter than 10 base pairs or the 
similarity over this aligned region is below 80% of identity, in the output file,
the forward and reverse reads are not aligned but concatenated, and the value of 
the `mode` attribute in the sequence header is set to `joined` instead of `alignment`.

### Remove unaligned sequence records

Unaligned sequences (:py`mode=joined`{.interpreted-text role="mod"})
cannot be used. The following command allows removing them from the
dataset:

```{bash}
#| output: false

obigrep -p 'annotations.mode != "join"' \
        results/wolf.fastq > results/wolf.ali.fastq
```

The `-p` requires a go like expression. `annotations.mode != "join"` means that
if the value of the `mode` annotation of a sequence is
different from `join`, the corresponding sequence record will be kept.

The first sequence record of `wolf.ali.fastq` can be obtained using the
following command line:

```{bash}
#| eval: false
#| output: false

head -n 4 results/wolf.ali.fastq
```

The folling piece of code appears on thew window of tour terminal.

```
@HELIUM_000100422_612GNAAXX:7:108:5640:3823#0/1 {"ali_dir":"left","ali_length":62,"mode":"alignment","pairing_mismatches":{"(T:26)->(G:13)":62,"(T:34)->(G:18)":48},"score":484,"score_norm":0.968,"seq_a_single":46,"seq_ab_match":60,"seq_b_single":46}
ccgcctcctttagataccccactatgcttagccctaaacacaagtaattaatataacaaaattgttcgccagagtactaccggcaatagcttaaaactcaaaggacttggcggtgctttatacccttctagaggagcctgttctaaggaggcgg
+
CCCCCCCBCCCCCCCCCCCCCCCCCCCCCCBCCCCCBCCCCCCC<CcCccbe[`F`accXV<TA\RYU\\ee_e[XZ[XEEEEEEEEEE?EEEEEEEEEEDEEEEEEECCCCCCCCCCCCCCCCCCCCCCCACCCCCACCCCCCCCCCCCCCCC
```

### Assign each sequence record to the corresponding sample/marker combination

Each sequence record is assigned to its corresponding sample and marker
using the data provided in a text file (here `wolf_diet_ngsfilter.txt`).
This text file contains one line per sample, with the name of the
experiment (several experiments can be included in the same file), the
name of the tags (for example: `aattaac` if the same tag has been used
on each extremity of the PCR products, or `aattaac:gaagtag` if the tags
were different), the sequence of the forward primer, the sequence of the
reverse primer, the letter `T` or `F` for sample identification using
the forward primer and tag only or using both primers and both tags,
respectively (see `obimultiplex` for details).

```{bash}
#| output: false

obimultiplex -t wolf_data/wolf_diet_ngsfilter.txt \
             -u results/unidentified.fastq \
             results/wolf.ali.fastq \
             > results/wolf.ali.assigned.fastq
```

This command creates two files:

-   `unidentified.fastq` containing all the sequence records that were
    not assigned to a sample/marker combination
-   `wolf.ali.assigned.fastq` containing all the sequence records that
    were properly assigned to a sample/marker combination

Note that each sequence record of the `wolf.ali.assigned.fastq` file
contains only the barcode sequence as the sequences of primers and tags
are removed by the `obimultiplex ` program. Information concerning the 
experiment, sample, primers and tags is added as attributes in the 
sequence header.

For instance, the first sequence record of `wolf.ali.assigned.fastq` is:

```
@HELIUM_000100422_612GNAAXX:7:108:5640:3823#0/1_sub[28..127] {"ali_dir":"left","ali_length":62,"direction":"direct","experiment":"wolf_diet","forward_match":"ttagataccccactatgc","forward_mismatches":0,"forward_primer":"ttagataccccactatgc","forward_tag":"gcctcct","mode":"alignment","pairing_mismatches":{"(T:26)->(G:13)":35,"(T:34)->(G:18)":21},"reverse_match":"tagaacaggctcctctag","reverse_mismatches":0,"reverse_primer":"tagaacaggctcctctag","reverse_tag":"gcctcct","sample":"29a_F260619","score":484,"score_norm":0.968,"seq_a_single":46,"seq_ab_match":60,"seq_b_single":46}
ttagccctaaacacaagtaattaatataacaaaattgttcgccagagtactaccggcaatagcttaaaactcaaaggacttggcggtgctttataccctt
+
CCCBCCCCCBCCCCCCC<CcCccbe[`F`accXV<TA\RYU\\ee_e[XZ[XEEEEEEEEEE?EEEEEEEEEEDEEEEEEECCCCCCCCCCCCCCCCCCC
```

### Dereplicate reads into uniq sequences

The same DNA molecule can be sequenced several times. In order to reduce
both file size and computations time, and to get easier interpretable
results, it is convenient to work with unique *sequences* instead of
*reads*. To *dereplicate* such *reads* into unique *sequences*, we use
the `obiuniq` command.

+-------------------------------------------------------------+
| Definition: Dereplicate reads into unique sequences         |
+-------------------------------------------------------------+
| 1.  compare all the reads in a data set to each other       |
| 2.  group strictly identical reads together                 |
| 3.  output the sequence for each group and its count in the |
|     original dataset (in this way, all duplicated reads are |
|     removed)                                                |
|                                                             |
| Definition adapted from @Seguritan2001-tg                   |
+-------------------------------------------------------------+

For dereplication, we use the `obiuniq ` command with the `-m sample`. The `-m sample` option is used
to keep the information of the samples of origin for each uniquesequence.

```{bash}
#| output: false

obiuniq -m sample \
        results/wolf.ali.assigned.fastq \
        > results/wolf.ali.assigned.uniq.fasta
```

Note that `obiuniq` returns a fasta file.

The first sequence record of `wolf.ali.assigned.uniq.fasta` is:

```
>HELIUM_000100422_612GNAAXX:7:93:6991:1942#0/1_sub[28..126] {"ali_dir":"left","ali_length":63,"count":1,"direction":"reverse","experiment":"wolf_diet","forward_match":"ttagataccccactatgc","forward_mismatches":0,"forward_primer":"ttagataccccactatgc","forward_tag":"gaatatc","merged_sample":{"26a_F040644":1},"mode":"alignment","pairing_mismatches":{"(A:10)->(G:34)":76,"(C:06)->(A:34)":58},"reverse_match":"tagaacaggctcctctag","reverse_mismatches":0,"reverse_primer":"tagaacaggctcctctag","reverse_tag":"gaatatc","score":730,"score_norm":0.968,"seq_a_single":45,"seq_ab_match":61,"seq_b_single":45}
ttagccctaaacataaacattcaataaacaagaatgttcgccagagaactactagcaaca
gcctgaaactcaaaggacttggcggtgctttatatccct
```

The run of `obiuniq` has
added two key=values entries in the header of the fasta sequence:

-   `"merged_sample":{"29a_F260619":1}`{.interpreted-text
    role="mod"}: this sequence have been found once in a single sample
    called **29a_F260619**
-   `"count":1` : the total count for this sequence is $1$

To keep only these two attributes, we can use the `obiannotate` command:

```{bash}
#| output: false

obiannotate -k count -k merged_sample \
  results/wolf.ali.assigned.uniq.fasta \
  > results/wolf.ali.assigned.simple.fasta
```

The first five sequence records of `wolf.ali.assigned.simple.fasta`
become:

```
>HELIUM_000100422_612GNAAXX:7:26:18930:11105#0/1_sub[28..127] {"count":1,"merged_sample":{"29a_F260619":1}}
ttagccctaaacacaagtaattaatataacaaaatwattcgcyagagtactacmggcaat
agctyaaarctcamagrwcttggcggtgctttataccctt
>HELIUM_000100422_612GNAAXX:7:58:5711:11399#0/1_sub[28..127] {"count":1,"merged_sample":{"29a_F260619":1}}
ttagccctaaacacaagtaattaatataacaaaattattcgccagagtwctaccgssaat
agcttaaaactcaaaggactgggcggtgctttataccctt
>HELIUM_000100422_612GNAAXX:7:100:15836:9304#0/1_sub[28..127] {"count":1,"merged_sample":{"29a_F260619":1}}
ttagccctaaacatagataattacacaaacaaaattgttcaccagagtactagcggcaac
agcttaaaactcaaaggacttggcggtgctttataccctt
>HELIUM_000100422_612GNAAXX:7:55:13242:9085#0/1_sub[28..126] {"count":4,"merged_sample":{"26a_F040644":4}}
ttagccctaaacataaacattcaataaacaagagtgttcgccagagtactactagcaaca
gcctgaaactcaaaggacttggcggtgctttacatccct
>HELIUM_000100422_612GNAAXX:7:86:8429:13723#0/1_sub[28..127] {"count":7,"merged_sample":{"15a_F730814":5,"29a_F260619":2}}
ttagccctaaacacaagtaattaatataacaaaattattcgccagagtactaccggcaat
agcttaaaactcaaaggactcggcggtgctttataccctt
```

### Denoise the sequence dataset

To have a set of sequences assigned to their corresponding samples does
not mean that all sequences are *biologically* meaningful i.e. some of
these sequences can contains PCR and/or sequencing errors, or chimeras.

#### Tag the sequences for PCR errors (sequence variants) {.unnumbered}

The `obiclean` program tags sequence variants as potential error generated during
PCR amplification. We ask it to keep the [head]{.title-ref} sequences (`-H` option) 
that are sequences which are not variants of another sequence with a count greater than 5% of their own count
(`-r 0.05` option).

```{bash}
#| output: false

obiclean -s sample -r 0.05 -H \
  results/wolf.ali.assigned.simple.fasta \
      > results/wolf.ali.assigned.simple.clean.fasta 
```

One of the sequence records of
`wolf.ali.assigned.simple.clean.fasta` is:

```
>HELIUM_000100422_612GNAAXX:7:66:4039:8016#0/1_sub[28..127] {"count":17,"merged_sample":{"13a_F730603":17},"obiclean_head":true,"obiclean_headcount":1,"obiclean_internalcount":0,"obi
clean_samplecount":1,"obiclean_singletoncount":0,"obiclean_status":{"13a_F730603":"h"},"obiclean_weight":{"13a_F730603":25}}
ctagccttaaacacaaatagttatgcaaacaaaactattcgccagagtactaccggcaac
agcccaaaactcaaaggacttggcggtgcttcacaccctt
```

To remove such sequences as much as possible, we first discard rare
sequences and then rsequence variants that likely correspond to
artifacts.



#### Get some statistics about sequence counts  {.unnumbered}

```{bash}
obicount results/wolf.ali.assigned.simple.clean.fasta
```

The dataset contains $4313$ sequences variant corresponding to 42452 sequence reads.
Most of the variants occur only a single time in the complete dataset and are usualy
named *singletons*

```{bash}
obigrep -p 'sequence.Count() == 1' results/wolf.ali.assigned.simple.clean.fasta \
    | obicount
```

In that dataset sigletons corresponds to $3511$ variants.

Using *R* and the `ROBIFastread` package able to read headers of the fasta files produced by *OBITools*,
we can get more complete statistics on the distribution of occurrencies.
 
```{r}
#| warning: false
library(ROBIFastread)
library(ggplot2)

seqs <- read_obifasta("results/wolf.ali.assigned.simple.clean.fasta",keys="count")

ggplot(data = seqs,  mapping=aes(x = count)) +
  geom_histogram(bins=100) +
  scale_y_sqrt() +
  scale_x_sqrt() +
  geom_vline(xintercept = 10, col="red", lty=2) +
  xlab("number of occurrencies of a variant") 
```

In a similar way it is also possible to plot the distribution of the sequence length.

```{r}
#| warning: false
ggplot(data = seqs,  mapping=aes(x = nchar(sequence))) +
  geom_histogram() +
  scale_y_log10() +
  geom_vline(xintercept = 80, col="red", lty=2) +
  xlab("sequence lengths in base pair")
```


#### Keep only the sequences having a count greater or equal to 10 and a length shorter than 80 bp  {.unnumbered}

Based on the previous observation, we set the cut-off for keeping
sequences for further analysis to a count of 10. To do this, we use the
`obigrep <scripts/obigrep>`{.interpreted-text role="doc"} command. The
`-p 'count>=10'` option means that the `python` expression
:py`count>=10`{.interpreted-text role="mod"} must be evaluated to
:py`True`{.interpreted-text role="mod"} for each sequence to be kept.
Based on previous knowledge we also remove sequences with a length
shorter than 80 bp (option -l) as we know that the amplified 12S-V5
barcode for vertebrates must have a length around 100bp.

```{bash}
#| output: false

obigrep -l 80 -p 'sequence.Count() >= 10' results/wolf.ali.assigned.simple.clean.fasta \
    > results/wolf.ali.assigned.simple.clean.c10.l80.fasta
```

The first sequence record of `results/wolf.ali.assigned.simple.clean.c10.l80.fasta` is:
 
```
>HELIUM_000100422_612GNAAXX:7:22:2603:18023#0/1_sub[28..127] {"count":12182,"merged_sample":{"15a_F730814":7559,"29a_F260619":4623},"obiclean_head":true,"obiclean_headcount":2,"obiclean_internalcount":0,"obiclean_samplecount":2,"obiclean_singletoncount":0,"obiclean_status":{"15a_F730814":"h","29a_F260619":"h"},"obiclean_weight":{"15a_F730814":9165,"29a_F260619":6275}}
ttagccctaaacacaagtaattaatataacaaaattattcgccagagtactaccggcaat
agcttaaaactcaaaggacttggcggtgctttataccctt
```

At that time in the data cleanning we have conserved : 

```{bash}
obicount results/wolf.ali.assigned.simple.clean.c10.l80.fasta
```

### Taxonomic assignment of sequences

Once denoising has been done, the next step in diet analysis is to
assign the barcodes to the corresponding species in order to get the
complete list of species associated to each sample.

Taxonomic assignment of sequences requires a reference database
compiling all possible species to be identified in the sample.
Assignment is then done based on sequence comparison between sample
sequences and reference sequences.

#### Download the taxonomy  {.unnumbered}

It is always possible to download the complete taxonomy from NCBI using the following commands.

```{bash}
#| output: false
mkdir TAXO
cd TAXO
curl http://ftp.ncbi.nih.gov/pub/taxonomy/taxdump.tar.gz \
   | tar -zxvf -
cd ..
```

For people have a low speed internet connection, a copy of the `taxdump.tar.gz` file is provided in the wolf_data directory.
The NCBI taxonomy is dayly updated, but the one provided here is ok for running this tutorial.

To build the TAXO directory from the provided `taxdump.tar.gz`, you need to execute the following commands

```{bash}
#| output: false
mkdir TAXO
cd TAXO
tar zxvf wolf_data/taxdump.tar.gz 
cd ..
```

#### Build a reference database  {.unnumbered}

One way to build the reference database is to use the `obipcr` program to simulate a PCR and extract all sequences from a general purpose DNA database such as genbank or EMBL that can be
amplified *in silico* by the two primers (here **TTAGATACCCCACTATGC** and **TAGAACAGGCTCCTCTAG**)
used for PCR amplification.

The two steps to build this reference database would then be

1.  Today, the easiest database to download is *Genbank*. But this will take you more than a day and occupy more than half a terabyte on your hard drive. In the `wolf_data` directory, a shell script called `download_gb.sh` is provided to perform this task. It requires that the programs `wget2` and `curl` are available   on your computer.

1.  Use `obipcr` to simulate amplification and build a reference database based on the putatively amplified barcodes and their recorded taxonomic information.

As these steps can take a long time (about a day for the download and an hour for the PCR), we already provide the reference database produced by the following commands so you can skip its construction. Note that as the Genbank and taxonomic database evolve frequently, if you run the following commands you may get different results.

##### Download the sequences  {.unnumbered}

```{bash}
#| eval: false
mkdir genbank
cd genbank
../wolf_data/install_gb.sh
cd ..
``` 

DO NOT RUN THIS COMMAND EXCEPT IF YOU ARE REALLY CONSIENT OF THE TIME AND DISK SPACE REQUIRED.


##### Use obipcr to simulate an in silico\` PCR  {.unnumbered}

```{bash}
#| eval: false
obipcr -t TAXO -e 3 -l 50 -L 150 \ 
       --forward TTAGATACCCCACTATGC \
       --reverse TAGAACAGGCTCCTCTAG \
       --no-order \
       genbank/Release-251/gb*.seq.gz
       > results/v05.pcr.fasta
```

Note that the primers must be in the same order both in
`wolf_diet_ngsfilter.txt` and in the `obipcr` command.
The part of the path indicating the *Genbank* release can change.
Please check in your genbank directory the exact name of your release.

##### Clean the database  {.unnumbered}

1.  filter sequences so that they have a good taxonomic description at
    the species, genus, and family levels
    (`obigrep` command command below).
2.  remove redundant sequences  (`obiuniq` command below).
3.  ensure that the dereplicated sequences have a taxid at the family
    level (`obigrep` command below).
4.  ensure that sequences each have a unique identification
    (`obiannotate` command below)

```{bash}
#| eval: false

obigrep -t TAXO \
          --require-rank species \
          --require-rank genus \
          --require-rank family \
          results/v05.ecopcr > results/v05_clean.fasta

obiuniq -c taxid \
        results/v05_clean.fasta \
        > results/v05_clean_uniq.fasta

obirefidx -t TAXO results/v05_clean_uniq.fasta \
        > results/v05_clean_uniq.indexed.fasta
```



::: warning
::: title
Warning
:::

From now on, for the sake of clarity, the following commands will use
the filenames of the files provided with the tutorial. If you decided to
run the last steps and use the files you have produced, you\'ll have to
use `results/v05_clean_uniq.indexed.fasta` instead of `wolf_data/db_v05_r117.indexed.fasta`.
:::

### Assign each sequence to a taxon

Once the reference database is built, taxonomic assignment can be
carried out using the `obitag` command.

```{bash}
#| output: false
obitag -t TAXO -R wolf_data/db_v05_r117.indexed.fasta \
       results/wolf.ali.assigned.simple.clean.c10.l80.fasta \
       > results/wolf.ali.assigned.simple.clean.c10.l80.taxo.fasta
```

The `obitag` adds several attributes in the sequence record header, among
them:

-   obitag_bestmatch=ACCESSION where ACCESSION is the id of hte sequence in
    the reference database that best aligns to the query sequence;
-   obitag_bestid=FLOAT where FLOAT\*100 is the percentage of identity
    between the best match sequence and the query sequence;
-   taxid=TAXID where TAXID is the final assignation of the sequence by
    `obitag`
-   scientific_name=NAME where NAME is the scientific name of the
    assigned taxid.

The first sequence record of `wolf.ali.assigned.simple.clean.c10.l80.taxo.fasta` is:

``` bash
>HELIUM_000100422_612GNAAXX:7:81:18704:12346#0/1_sub[28..126] {"count":88,"merged_sample":{"26a_F040644":88},"obiclean_head":true,"obiclean_headcount":1,"obiclean_internalcount":0,"obiclean_samplecount":1,"obiclean_singletoncount":0,"obiclean_status":{"26a_F040644":"h"},"obiclean_weight":{"26a_F040644":208},"obitag_bestid":0.9207920792079208,"obitag_bestmatch":"AY769263","obitag_difference":8,"obitag_match_count":1,"obitag_rank":"clade","scientific_name":"Boreoeutheria","taxid":1437010}
ttagccctaaacataaacattcaataaacaagaatgttcgccagaggactactagcaata
gcttaaaactcaaaggacttggcggtgctttatatccct
```

### Generate the final result table

Some unuseful attributes can be removed at this stage.

- obiclean_head
- obiclean_headcount
- obiclean_internalcount
- obiclean_samplecount
- obiclean_singletoncount

```{bash}
#| output: false
obiannotate  --delete-tag=obiclean_head \
             --delete-tag=obiclean_headcount \
             --delete-tag=obiclean_internalcount \
             --delete-tag=obiclean_samplecount \
             --delete-tag=obiclean_singletoncount \
  results/wolf.ali.assigned.simple.clean.c10.l80.taxo.fasta \
  > results/wolf.ali.assigned.simple.clean.c10.l80.taxo.ann.fasta
```

The first sequence record of
`wolf.ali.assigned.simple.c10.l80.clean.taxo.ann.fasta` is then:
 
``` 
>HELIUM_000100422_612GNAAXX:7:84:16335:5083#0/1_sub[28..126] {"count":96,"merged_sample":{"26a_F040644":11,"29a_F260619":85},"obiclean_status":{"26a_F040644":"s","29a_F260619":"h"},"obiclean_weight":{"26a_F040644":14,"29a_F260619":110},"obitag_bestid":0.9595959595959596,"obitag_bestmatch":"AC187326","obitag_difference":4,"obitag_match_count":1,"obitag_rank":"subspecies","scientific_name":"Canis lupus familiaris","taxid":9615}
ttagccctaaacataagctattccataacaaaataattcgccagagaactactagcaaca
gattaaacctcaaaggacttggcagtgctttatacccct
```

### Looking at the data in R


```{r}
library(ROBIFastread)
library(vegan)
library(magrittr)
 

diet_data <- read_obifasta("results/wolf.ali.assigned.simple.clean.c10.l80.taxo.fasta") 
diet_data %<>% extract_features("obitag_bestmatch","obitag_rank","scientific_name",'taxid')

diet_tab <- extract_readcount(diet_data,key="obiclean_weight")
diet_tab
```

This file contains 26 sequences. You can deduce the diet of each sample: 

:   -   13a_F730603: Cervus elaphus
    -   15a_F730814: Capreolus capreolus
    -   26a_F040644: Marmota sp. (according to the location, it is
        Marmota marmota)
    -   29a_F260619: Capreolus capreolus

Note that we also obtained a few wolf sequences although a wolf-blocking
oligonucleotide was used.
